---
title: Ted Petrou follow up analysis
author: Caleb Moses
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
    prettydoc::html_pretty:
      theme: tactile
highlight: github
---

# Introduction

This notebook investigates claims by Twitterer, data scientist and author Ted Petrou.

```{r import-libraries, message = FALSE, echo = FALSE}
library(tidyverse)
library(lubridate)
library(tweetrmd)
library(prophet)
library(here)

# Set stan cores
options(mc.cores = 8)

# Set default ggplot theme
theme_set(theme_minimal() + theme(plot.title = element_text(hjust = 0.5)))

# Set knitr options
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", warning = FALSE, message = FALSE, fig.width = 12)

```

The offending tweet:

```{r show-tweet, echo = FALSE, fig.align = "center"}

include_tweet("https://twitter.com/TedPetrou/status/1358441124894015490")

```

In the thread, Petrou goes on to predict that notes a decrease in mortality among
the aged 80+ cohort in 2020 (due to the March and August covid-19 lockdowns), and
predicts that this decrease will be wiped out by an increase in mortality.

## Reproducing the original plot

He does so on the basis of the plot in his first tweet (shown above). We begin
by reproducing his plot to ensure we are both using the same data.

```{r reproduce-petrou-plot, message = FALSE}

deaths_over80 <- read_csv(here('data/nz_weekly_deaths_by_age.csv')) %>%
    rename(Age = series_name, Date = parameter, Deaths = value) %>%
    select(Age, Date, Deaths) %>%
    filter(Age == '80 and over')

weekly_deaths_over80 <- deaths_over80 %>%
    mutate(Date = floor_date(Date, unit = "month")) %>%
    group_by(Age, Date) %>%
    summarise(Deaths = mean(Deaths))

deaths_2021 <- tibble(year = 2021, month = 2:12, day = 1) %>%
    mutate(Date = as.Date(sprintf("%s-%s-%s", year, month, day)),
           Deaths = 0, Age = '80 and over') %>%
    select(Age, Date, Deaths)

weekly_deaths_over80 <- bind_rows(weekly_deaths_over80, deaths_2021)

weekly_deaths_over80 %>%
    filter(year(Date) >= 2019) %>%
    mutate(Year = as.factor(year(Date)),
           Month = month(Date,label=TRUE,abbr=TRUE)) %>%
    ggplot(aes(x = Month, y = Deaths, fill = Year)) +
    geom_bar(stat = 'identity', position = 'dodge') +
    ggtitle("New Zealand deaths per month aged 80+")

```

I used data from the stats NZ covid 19 data portal, available here:
https://www.stats.govt.nz/experimental/covid-19-data-portal

## Show observations from 2011

One of the first things people looked at was the observations preceding 2019, which 
begins in 2011. We show this plot here:

```{r observations-from-2011}

weekly_deaths_over80 %>%
    mutate(Year = as.factor(year(Date)),
           Month = month(Date,label=TRUE,abbr=TRUE)) %>%
    ggplot(aes(x = Month, y = Deaths, fill = Year)) +
    geom_bar(stat = 'identity', position = 'dodge') +
    ggtitle("New Zealand deaths per month aged 80+")

```

Some have said that this shows the 2021 values are within the typical variation
of the data, which I would say is not strictly clear from this graph because
there is a long term trend and seasonality, which is visible when you view the
full time series. 

## Show full time series

I show the full time series (with all observations) here:

```{r full-time-series}

deaths_over80 %>%
    select(Age, Date, Deaths) %>%
    filter(Age == '80 and over') %>%
    ggplot(aes(x = Date, y = Deaths)) +
    geom_point() +
    expand_limits(x = min(deaths_over80$Date), y = 0)

```

# Modelling long term + seasonal trend for deaths over 80

Ideally you would remove the long term + seasonal trend before you drawing a
conclusion about whether the recent observations are ouside of the norm. We do this next:

```{r fit-prophet}

# Fit a model to the pre-2020 data
m <- deaths_over80 %>%
    filter(year(Date) < 2020) %>%
    rename(ds = Date, y = Deaths) %>%
    prophet(mcmc.samples = 2000,
            weekly.seasonality = FALSE,
            daily.seasonality = FALSE)

future <- make_future_dataframe(m, periods = 500)

forecast <- predict(m, future)

deaths_over80 %>%
    mutate(Predicted = if_else(year(Date) < 2020, 'Fit to model', 'Predicted')) %>%
    rename(ds = Date, ytrue = Deaths) %>%
    full_join(forecast) %>%
    mutate(yhat = if_else(year(ds) >= 2020, NA_real_, yhat)) %>%
    arrange(ds) %>%
    filter(!is.na(Predicted)) %>%
    ggplot(aes(x = ds, y = ytrue)) +
    geom_ribbon(aes(ymin = yhat_lower,
                    ymax = yhat_upper,
                    fill = Predicted),
                alpha = 0.5) +
    geom_point() +
    geom_line(aes(y = trend + additive_terms)) +
    xlab("Date") +
    ylab("Number of deaths") +
    ggtitle("Weekly deaths for people aged 80+ in New Zealand")

```

We fit a model to the pre-2020 data and predict the 2020 values to provide
a baseline for how many deaths were expected in each month.

## Show long term + seasonal trend

We can see the long term + seasonal trend here.

```{r}

prophet_plot_components(m, forecast)

```

## Calculate excess mortality post-2019

Here we calculate the post-2019 excess mortality. This plot shows the total excess 
deaths for the aged 80+ population which cannot be explained by the model.

```{r}

samples <- predictive_samples(m, future)

trend <- as_tibble(samples$trend) %>%
    gather(V1:V4000, key = 'chain', value = 'trend') %>%
    group_by(chain) %>%
    mutate(timestep = 1:n(),
           chain = as.integer(str_remove(chain, "V")))

yhat <- as_tibble(samples$yhat) %>%
    gather(V1:V4000, key = 'chain', value = 'yhat') %>%
    group_by(chain) %>%
    mutate(timestep = 1:n(),
           chain = as.integer(str_remove(chain, "V")))

sample_data <- full_join(trend, yhat) %>%
    select(chain, timestep, trend, yhat)

sample_data <- deaths_over80 %>%
    rename(ytrue = Deaths) %>%
    mutate(timestep = 1:nrow(.)) %>%
    full_join(sample_data) %>%
    select(Age, Date, chain, timestep, ytrue, trend, yhat)

sample_data %>%
    filter(Date >= as.Date("2020-01-01")) %>%
    group_by(chain) %>%
    mutate(cum_delta = cumsum(yhat - ytrue)) %>%
    group_by(Date) %>%
    summarise(cum_delta_lower = quantile(cum_delta, 0.05),
              cum_delta_upper = quantile(cum_delta, 0.95),
              cum_delta = mean(cum_delta)) %>%
    ggplot(aes(x = Date, ymin = cum_delta_lower, ymax = cum_delta_upper, y = cum_delta)) +
    geom_line() +
    geom_ribbon(fill = 'steelblue', alpha = 0.5) +
    ggtitle("Cumulative excess mortality for people aged 80+ since Jan 2020") +
    ylab("Excess mortality")

```

The plot shows there are between 500 and 1500 people aged 80+ who would have died in 2020
but are still alive.

Additionally, the plot shows no evidence of an upswing back towards zero. In other words, the
data so far does not provide evidence that we are experiencing above average deaths to 
compensate for those saved last year.

## Comparing 2020 deaths to the model

Lastly, we can specifically compare the 2020 values to the model. We can see the model is a
poor fit, but this is okay because we the model represents the deaths we would have seen
in a typical year.

```{r}

deaths_over80 %>%
    rename(ds = Date, ytrue = Deaths) %>%
    full_join(forecast) %>%
    filter(year(ds) >= 2020) %>%
    filter(ds < as.Date('2021-02-01')) %>%
    ggplot(aes(x = ds, y = ytrue)) +
    geom_ribbon(aes(ymin = yhat_lower,
                    ymax = yhat_upper),
                fill = "steelblue", alpha = 0.5) +
    geom_point() +
    geom_line(aes(y = trend + additive_terms)) +
    xlab("Date") +
    ylab("Deaths") +
    ggtitle("Weekly deaths for people aged 80+ from 2020 to present")

```
